Localization properties, observable character and computational reliability of the linear response in open systems

### 2018-06-06 AnalyticStructure_main.tex
intriduction, first versions

\section{Introduction}

The subject of the present investigation concerns with the study of isolated molecular-like systems subjected to a time-dependent perturbation. % in the linear response regime.
This problem can be tackled within the contest of the time-dependent density functional theory (TD-DFT) \cite{runge1984,onida2002}, that allows one to establish the time-dependent 
Kohn-Sham (KS) equations which determine the evolution of the one-particle properties of the system with respect to time.

We focus our attention to the linear response regime (LR) in which the theory can be conveniently expressed in the frequency domain and the time-dependent KS equation are reformulated 
as a quantum Liouville equation for the response density operator. Different computational approaches have been developed in literature \cite{baroni2008,hubener2014,brabec2015}, principally 
with the aim of producing efficient algorithms for the evaluation of the physical observables. These methods differ both for the procedure adopted to solve the Liouville equation 
(explicit diagonalization, self-consistent Sternheimer equation, Lanczos chain-like procedure, etc.) and both for the choice of the basis set employed to perform the calculations. 
Nonetheless, in all the cases one is a dealing with a finite basis set that, unavoidably, gives rise to a limited computational domain. 
It is then meaningful to try to understand under 
which extent the computational-induced truncation of the natural open boundaries condition could produce observable effects on the physical quantities computed in the LR.

A preliminary understanding of this problem can be achieved thanks to the results reported in \cite{boffi2016}. The authors of this paper discuss the effects of a finite basis set on the 
occupied and virtual orbitals of a molecular system and establish a clear correspondence between the orbital localization character, and the independence of its properties (\emph{e.g.} energy) 
with respect to the basis set. It is shown that (localized) orbitals with a bound-state behavior in the long range constitute a genuine discrete set and a finite basis is 
able to provide a reliable description of these states if the sampled domain contains their natural sizes. On the contrary, unbound (delocalized) orbitals form a continuum of states in the 
true system with open boundaries. For these class of objects the presence of a finite computational domain has a deeper impact since introduces a fictitious discretization of the energy levels. 
Furthermore, the energies of the unbound orbitals are not stable with respect to a change of the size of the computational domain and approach to zero as long as it is increased.

In this paper we aim to reach a deeper understanding of the possible effects of a finite computational basis set on the LR regime in open systems. To achieve this task we try to
reformulate the relation between localization properties and basis independence for the main actors of the LR formalism. We present a discussion based on a double level analysis. First, 
we propose a localization argument directly based on the inspection of the response density operator. Then, we move to a deeper level and investigate the properties of the susceptibility 
functional, which represents the fundamental object in the contest of the linear response. Pursuing this approach we are naturally led to investigate the localization properties of excitation 
operators, that in turn emerge as the resonant channel of the spectral decomposition of the susceptibility. In this way we establish a direct relation between the concept of basis independence 
and the analytic structure of the linear response susceptibility.

>>>>

\section{Introduction}
Since the days of their foundations, disciplines like Computational Physics or Quantum Chemistry always had to deal with the problem of the 
computational reliability of the results. More specifically, a computational result may be considered as reliable for various reasons.
It might be, for example, in agreement with some previously existing data that the employed computational model is susceptible to reproduce.
The model itself can in this way be classified as ``accurate''; it may then be employed to extract, \textit{in silico},
other quantities that might be then used to \emph{predict} or \emph{interpret} experimental data.
Another concept of reliability is the concept of  ``precision'', which is more focused to the actual results
that a given computational model provides for a specific set of quantities, regardless of their potential agreement with 
experimental data. The aim of precise computational approach is to reduce the \emph{computational} uncertanties
of quantities extracted out of a well-defined model, and provide reference results, to which other computer codes
employing the same model can be compared.

A recent, remarkable exemple of precision-driven initiative is provided by the DeltaCodes project~\cite{deltaTest2016},
where a grassroot community of code-owners struggled to extract well-defined computational results in the context of 
Density Functional Theory (DFT) calculations. Ground-State quantum mechanical quantities as Lattice constants and bulk modulus of 
elementary crystals extracted from different codes were compared with each other, with the objective of reducing the uncertanties 
on such data, and in some sense, to define a \emph{calibration} procedure for a Solid-State Physics computer code. 
Any other code able to extract the same quantites with the same approximations might be compared with the 
results of the community in order to assess its computational reliability.

For a theoretical approach that is based on numerical calculations, for which no analytic reference solution exists, 
reducing the computational uncertainty is the only possible way to shed light on the predictivity power of the
model. In other terms, the ``accuracy'' of a result with respect to experimental data may be reliably quantified
only when the computational uncertainty is guaranteed to be significantly lower than the found discrepancy.
For communities like DFT, such a ``calibration'' of computer codes among each other is a discipline that is still in its infancy;
the DeltaCodes initiative and the foundations of DFT are separated in time by more than 50 years, and the computational
model employed in the comparison (the Perdew-Burke-Erzerhof functional~\cite{PBE}) has been introduced more than 20 years ago.

The problem is even more stringent for calculations that refer to quantities \emph{beyond} ground state, like
for instance the time-dependent density functional theory (TDDFT) \cite{casida1995,runge1984,onida2002},
from which quantities like optical excitations or polarizability tensors
of a given system might be studied and put in relation with experimental data.
Here also, a plethora of computational approaches exist, that employ different numerical formalism and
basis sets, and it will be of great importance to identify quantities that can be ``measured'' in a given model, to assess
and quantify their computational uncertainty.

Moreover, wheareas for Ground-State calculations the quantities are relatively easy to identify and extract,
TD-DFT and more generally, linear-response calculation present a further difficulty, which is the problem
of \emph{numerical} convergence. In other terms, beyond ground state, 
the absence of tools like the variational theorem imposes us to focus on \emph{well-defined} quantites,
and to study the conditions under which numerical convergence of the results can be achieved.

In this paper, we focus our attention to Linear-Response treatments of \emph{molecular} systems, 
in particular referring to LR-TDDFT calculation. We will investigate the impact that 
\emph{open} (i.e. Isolated) Boundary Conditions (BC) that have to be imposed to the system will have on the numerical convergence of LR quantites.
In particular, we will discuss the interplay between the numerical convergence of LR quantities and the \emph{locality}
of the related objects.
In order to do this, we perform analytic considerations on the equations beyond LR treatment,
which will help us identifying how to distinguish -- \textit{a priori} -- quantities that are susceptible to be 
compared among different treatments, as they are \emph{intrinsically} associated to localized objects,
from quantities that \emph{explicitly} depend of the the employed computational treatment.

Our consideration will allso reveal useful for the understanding of the \emph{analytic structure} of LR quantities like
the linear susceptibility of the system (i.e. the \emph{reducible} polarizability).
Altough we refer, as anticipated, to TDDFT models, such considerations might also be transposed to Many-Body Perturbation Theory
approaches, in the context Quasi-Particle equations.

% The subject of the present investigation concerns with the study of isolated molecular-like systems subjected to a time-dependent perturbation. % in the linear response regime.
% This problem can be tackled within the contest of the time-dependent density functional theory (TD-DFT) \cite{runge1984,onida2002}, that allows one to establish the time-dependent 
% Kohn-Sham (KS) equations which determine the evolution of the one-particle properties of the system with respect to time. 
% 
% We focus our attention to the linear response regime (LR) in which the theory can be conveniently expressed in the frequency domain and the time-dependent KS equation are reformulated 
% as a quantum Liouville equation for the response density operator. Different computational approaches have been developed in literature \cite{baroni2008,hubener2014,brabec2015}, principally 
% with the aim of producing efficient algorithms for the evaluation of the physical observables. These methods differ both for the procedure adopted to solve the Liouville equation 
% (explicit diagonalization, self-consistent Sternheimer equation, Lanczos chain-like procedure, etc.) and both for the choice of the basis set employed to perform the calculations. 
% Nonetheless, in all the cases one is a dealing with a finite basis set that, unavoidably, gives rise to a limited computational domain. 
% It is then meaningful to try to understand under 
% which extent the computational-induced truncation of the natural open boundaries condition could produce observable effects on the physical quantities computed in the LR. 
% 
% A preliminary understanding of this problem can be achieved thanks to the results reported in \cite{boffi2016}. The authors of this paper discuss the effects of a finite basis set on the 
% occupied and virtual orbitals of a molecular system and establish a clear correspondence between the orbital localization character, and the independence of its properties (\emph{e.g.} energy) 
% with respect to the basis set. It is shown that (localized) orbitals with a bound-state behavior in the long range constitute a genuine discrete set and a finite basis is 
% able to provide a reliable description of these states if the sampled domain contains their natural sizes. On the contrary, unbound (delocalized) orbitals form a continuum of states in the 
% true system with open boundaries. For these class of objects the presence of a finite computational domain has a deeper impact since introduces a fictitious discretization of the energy levels. 
% Furthermore, the energies of the unbound orbitals are not stable with respect to a change of the size of the computational domain and approach to zero as long as it is increased.  
% 
% In this paper we aim to reach a deeper understanding of the possible effects of a finite computational basis set on the LR regime in open systems. To achieve this task we try to
% reformulate the relation between localization properties and basis independence for the main actors of the LR formalism. We present a discussion based on a double level analysis. First, 
% we propose a localization argument directly based on the inspection of the response density operator. Then, we move to a deeper level and investigate the properties of the susceptibility 
% functional, which represents the fundamental object in the contest of the linear response. Pursuing this approach we are naturally led to investigate the localization properties of excitation 
% operators, that in turn emerge as the resonant channel of the spectral decomposition of the susceptibility. In this way we establish a direct relation between the concept of basis independence 
% and the analytic structure of the linear response susceptibility.
